LCS Tool — An Algorithmic Introduction to Lagrangian Coherent Structures

### 2014-02-04 main.tex
Revisions to abstract, introduction and Bickley jet section

\title{An Algorithmic Introduction to Lagrangian Coherent Structures}

@@

\abstract{
We give an applied introduction to Lagrangian coherent structure (LCS) detection. We review LCS theory that is based on variational calculus, present algorithms conceived in accordance with the variational theory and present a MATLAB computational engine called LCS Tool that enables variational LCS calculations and detection. We use LCS Tool to analyse three flow examples: a double gyre, a Bickley jet and an ocean dataset.
}

@@

LCSs are used in fluid dynamics to capture essential flow features and thereby simplify the description of complex fluid flows. More specifically, we analyze two dimensional finite time turbulent flows to detect LCSs. LCSs identify transport barriers, repelling and attracting structures and vortex boundaries. Application areas for LCSs are oceanic and atmospheric flows, biological applications\parencite{wilson09:_lagran_reynol,tallapragada11:_lagran}, aeronautics\parencite{tang10:_accur_lagran_hong_kong_inter_airpor} and mechanical systems\parencite{hadjighasem13:_detec_kam}.

The finite-time Lyapunov exponent (FTLE) gives one widely-used mathematical definition for LCSs. Although the FTLE is widely used, it is possible to find examples where it fails to identify LCSs correctly\parencite{haller11:_lagran_coher_struc,norgard12:_secon_lagran_coher_struc}. To remedy these problems, variational calculus applied to material curves in a fluid provides a rigorous approach for LCS detection\parencite{haller11:_lagran_coher_struc,farazmand12:_comput_lagran,haller12:_geodes_theor_trans_barrier_two_dimen_flows,farazmand13:_attrac_lagran,haller13:_coher_lagran}. This article presents an algorithmic introduction to variational LCSs supported by a computational engine called LCS Tool\footnote{LCS Tool is available at: \url{https://github.com/jeixav/LCS-Tool}.}.

LCS Tool is a library of MATLAB functions that perform computations to identify LCSs in two dimensional finite time flows. LCS Tool aims to help scientists perform variational LCS calculations. Other LCS software has been developed before; the list includes:
\begin{itemize}
\item \emph{MANGEN}\parencite{lekien03:_time}, calculates stability manifolds adaptively in two dimensional finite time velocity fields. Includes a graphical user interface and uses MPI for parallel calculations.
\item \emph{LCS MATLAB Kit}\parencite{dabiri09:_lmk} calculates the FTLE from velocity datasets. Includes a graphical user interface.
\item \emph{Newman}\parencite{toit10:_trans} calculates the FTLE in $N$ dimensions. Assists ridge extraction of FTLE fields. Supports analytic and dataset defined velocity definitions.
\item \emph{FlowVC}\parencite{shadden10:_flowvc} is a general purpose LCS platform for two and three dimensional datasets. Parallel calculations are supported with Open\-MP, CUDA and OpenCL.
\end{itemize}
These packages are based on the FTLE; the focus of LCS Tool and this article is to encourage and facilitate variational LCS analysis.

@@

The first step is to set the position of Poincare sections as straight lines in regions where closed lambda-lines are expected. The Poincare section must be oriented such that the first endpoint is close to the centre of the expected eddy and the second endpoint is outside the expected eddy. There is no definite procedure for deciding the position of Poincare sections. Possibilities include an examination of the finite-time Lyapunov exponent field of the flow, or some prior knowledge of where coherent eddies are be expected. Additionally, the number of lambda-lines launched from the Poincare section, \lstinline!poincareSection.numPoints!, must be defined. A good default is 100.

@@

\section{Examples}
In this section we present examples showing how to use LCS Tool to obtain LCSs in different flows, namely a double gyre, a jet, and an oceanic geostrophic flow. All three examples are available as demo files in the demo folder of LCS Tool and the user can follow the computation of LCS step by step.

@@

\lstinputlisting[caption={LCS Tool commands for strainline and stretchline LCSs},label=lst:double_gyre_lambda_strain_stretch_lcs]{listings/double_gyre/lambda_strain_stretch_lcs.m}

@@

Figure~\ref{fig:Bickley_jet_FTLE} shows the overall appearance of the Bickley jet and in Figure~\ref{fig:Bickley_jet_lambda_LCS_full} lambda-line LCSs are shown. The latter figure demonstrates a poor result since the LCSs change shape significantly from one eddy to the next. To address this problem we could try to adjust numerical parameters such as the resolution, but with our computer hardware memory availability was reached before achieving convergence. To circumvent this problem, we analyse lambda-line closed orbits of one vortex for a short timespan.

\begin{figure}
  \centering
  \includegraphics[width=.85\textwidth]{graphics/bickley_jet/ftle_overview}
  \caption{Bickley jet FTLE}
  \label{fig:Bickley_jet_FTLE}
\end{figure}

@@

The script used to analyse lambda-line LCS is given in Listing~\ref{lst:Bickley jet lambda lcs subdomain}. The script produces the data shown in Figure~\ref{fig:Bickley jet lambda lcs convergence}. Parameter values used may be read from the script. It is worth pointing out that in order to obtain closed orbits, $\lambda = 0.995$; with $\lambda = 1$ no closed orbits are found.

\lstinputlisting[caption={Bickley jet script for lambda-line LCSs},label=lst:Bickley jet lambda lcs subdomain]{listings/bickley_jet/lambda_lcs_convergence_subdomain.m}

@@

To calculate strainline and stretchline LCSs we set the timespan to $t \in [0,2L_x/U]$. The script used to detect strainlines is given in Listing~\ref{lst:bickley_jet_strain_lcs} and the results are shown in Figure~\ref{fig:Bickley_jet_strain_lcs}. Stretchline LCSs are shown in Figure~\ref{fig:Bickley_jet_stretch_lcs}.

\lstinputlisting[caption={LCS Tool commands for Bickley jet strainline LCSs},label=lst:bickley_jet_strain_lcs]{listings/bickley_jet/strain_lcs.m}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/ocean_dataset/hyperbolic_shearline_lcs_details_strainline}
\end{center}
\caption{Lambda-line LCS (green) and strainline LCS (red) of the ocean data set. Background gray-scale is the finite-time Lyapunov exponent. Dashed green lines are Poincare sections. White dots indicate $\lambda_2$ local maxima used for strainline integration initial positions.}
\label{f:ocean dataset hyperbolic shear lcs details strainline}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/ocean_dataset/hyperbolic_shearline_lcs_details_stretchline}
\end{center}
\caption{Lambda-line LCS (green) and stretchline LCS (blue) of the ocean data set. Background gray-scale is the finite-time Lyapunov exponent. Dashed green lines are Poincare sections. White dots indicate $\lambda_1$ local minima used for stretchline integration initial positions.}
\label{f:ocean dataset hyperbolic shear lcs details stretchline}
\end{figure}

@@

Algorithms to identify LCSs using variational principles have been presented, and a software library, LCS Tool, that implements these algorithms has been demonstrated on geophysical flow models and a dataset. The software library provides functions to enable exploring variational LCS methods without requiring detailed knowledge of variational LCS theory. These functions make it possible to write compact programs to identify and visualise LCSs.

@@

% Reduce number of user inputs, for example, automated closed orbit detection, which is especially important for flows with a large number of vortices possible.

>>>>

\title{LCS Tool — An Algorithmic Introduction to Lagrangian Coherent 
Structures}

@@

\abstract{
We give an applied introduction to Lagrangian coherent structures (LCSs) using a computational engine we have programmed called LCS Tool. First, we review LCS theory that is based on variational calculus; this introduces elliptic and hyperbolic variational LCSs. Then we turn our attention to algorithms for computations conceived in accordance with the variational LCS theory. These algorithms are the foundation of LCS Tool. We conclude by explaining, with demonstration scripts, how to use LCS Tool to analyse three flow examples: a double gyre, a Bickley jet and an ocean dataset.
}

@@

LCSs are used in fluid dynamics to capture essential flow features and thereby simplify the description of complex fluid flows. Here we focus on analysing two dimensional finite time turbulent flows. LCSs identify transport barriers, repelling and attracting structures and vortex boundaries. Application areas for LCSs include oceanic and atmospheric flows, biological applications\parencite{wilson09:_lagran_reynol,tallapragada11:_lagran}, aeronautics\parencite{tang10:_accur_lagran_hong_kong_inter_airpor} and mechanical systems\parencite{hadjighasem13:_detec_kam}.

The finite time Lyapunov exponent (FTLE) is one widely-used mathematical definition for LCSs. Despite its widespread use, examples where the FTLE fails to identify LCSs correctly are known\parencite{haller11:_lagran_coher_struc,norgard12:_secon_lagran_coher_struc}. To remedy these problems, variational calculus has been applied to material curve properties in the context of fluid mechanics. Thus we obtain a rigorous approach to LCS detection\parencite{haller11:_lagran_coher_struc,farazmand12:_comput_lagran,haller12:_geodes_theor_trans_barrier_two_dimen_flows,farazmand13:_attrac_lagran,haller13:_coher_lagran}. This article presents an algorithmic introduction to variational LCSs supported by a computational engine called LCS Tool\footnote{LCS Tool is available at: \url{https://github.com/jeixav/LCS-Tool}.}.

LCS Tool is a library of MATLAB functions that perform computations to identify LCSs in two dimensional finite time flows. LCS Tool aims to help scientists perform variational LCS calculations. The examples we present have been programmed into demonstration scripts. We encourage readers to download and execute these scripts after reading this article.

Other LCS software has been developed before; the list includes:
\begin{itemize}
\item \emph{MANGEN}\parencite{lekien03:_time}, calculates stability manifolds adaptively in two dimensional finite time velocity fields. Includes a graphical user interface and uses MPI for parallel calculations.
\item \emph{LCS MATLAB Kit}\parencite{dabiri09:_lmk} calculates the FTLE from velocity datasets. Includes a graphical user interface.
\item \emph{Newman}\parencite{toit10:_trans} calculates the FTLE in $N$ dimensions. Assists ridge extraction of FTLE fields. Supports analytic and dataset defined velocity definitions.
\item \emph{FlowVC}\parencite{shadden10:_flowvc} is a general purpose LCS platform for two and three dimensional datasets. Parallel calculations are supported with Open\-MP, CUDA and OpenCL.
\end{itemize}
These packages are based on the FTLE; the focus of LCS Tool and this article is to encourage and facilitate variational LCS analysis.

@@

The first step is to set the position of Poincare sections as straight lines in regions where closed lambda-lines are expected. The Poincare section must be oriented such that the first endpoint is close to the centre of the expected eddy and the second endpoint is outside the expected eddy. There is no definite procedure for deciding the position of Poincare sections. Possibilities include an examination of the finite time Lyapunov exponent field of the flow, or some prior knowledge of where coherent eddies are be expected. Additionally, the number of lambda-lines launched from the Poincare section, \lstinline!poincareSection.numPoints!, must be defined. A good default is 100.

@@

\section{Examples}
In this section we present examples showing how to use LCS Tool to obtain LCSs in different flows, namely a double gyre, a jet, and an oceanic geostrophic flow. All three examples are available as demonstration files in the demo folder of LCS Tool. By executing these scripts, readers can follow LCS Tool computations stepwise.

@@

\lstinputlisting[caption={LCS Tool commands for double gyre strainline and stretchline LCSs. Condensed version of the LCS Tool script \lstinline!demo/double_gyre/hyperbolic_shear_lcs.m! },label=lst:double_gyre_lambda_strain_stretch_lcs]{listings/double_gyre/lambda_strain_stretch_lcs.m}

@@

Figure~\ref{fig:Bickley_jet_lambda_LCS_full} shows lambda-line LCS of the Bickley jet with the FTLE in the background. The figure demonstrates a poor result since the LCSs change shape significantly between eddies. To address this problem we could adjust numerical parameters such as the resolution, but instead we will analyse lambda-line closed orbits of one vortex for a short timespan.

@@

Figure ~\ref{fig:Bickley jet lambda lcs convergence} shows convergence of the lambda-line elliptic LCSs. The LCS Tool script used to produce that figure are based on the script \lstinline!demo/bickley_jet/hyperbolic_shear_lcs.m! which is given in  Listing~\ref{lst:Bickley jet hyperbolic_shear_lcs subset}. It is worth pointing out that in order to obtain closed orbits, $\lambda = 0.995$; with $\lambda = 1$ no closed orbits exist.

\lstinputlisting[caption={Bickley jet script for hyperbolic and shear LCSs. Subset of \lstinline!demo/bickley_jet/hyperbolic_shear_lcs.m!},label=lst:Bickley jet hyperbolic_shear_lcs subset]{listings/bickley_jet/hyperbolic_shear_lcs_subset.m}

@@

Strainline LCSs are shown in Figure~\ref{fig:Bickley_jet_strain_lcs} and stretchline LCSs are shown in Figure~\ref{fig:Bickley_jet_stretch_lcs}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/ocean_dataset/hyperbolic_shearline_lcs_details_strainline}
\end{center}
\caption{Lambda-line LCSs (green) and strainline LCSs (red) of the ocean data set. Background gray-scale is the finite time Lyapunov exponent. Dashed green lines are Poincare sections. White dots indicate $\lambda_2$ local maxima used for strainline integration initial positions.}
\label{f:ocean dataset hyperbolic shear lcs details strainline}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/ocean_dataset/hyperbolic_shearline_lcs_details_stretchline}
\end{center}
\caption{Lambda-line LCSs (green) and stretchline LCSs (blue) of the ocean data set. Background gray-scale is the finite time Lyapunov exponent. Dashed green lines are Poincare sections. White dots indicate $\lambda_1$ local minima used for stretchline integration initial positions.}
\label{f:ocean dataset hyperbolic shear lcs details stretchline}
\end{figure}

@@

Algorithms to identify LCSs using variational principles have been presented, and a software library, LCS Tool, that implements these algorithms has been introduced and demonstrated on analytic flow models and a geophysical dataset. The software library provides functions to enable exploring variational LCS methods without requiring detailed knowledge of variational LCS theory. These functions make it possible to write compact programs to identify and visualise LCSs. Since LCS Tool is programmed in MATLAB, testing new methods and visualising their results can be done quite quickly.

@@
